---
title: "Getting and Cleaning Data Course Project Codebook"
author: "Xiaoning Wang"
date: "March 13, 2015"
output: html_document
---

### Preparation

In this course project, _dplyr_ library is used to clean and tidy the data set.
```{r load-library}
suppressMessages(library(dplyr))
setwd("~/Dropbox/Programming/Coursera/getdata-012/getdata-012-project")
```

Next, check if the data file exists in the working directory, download from source URL if necessary.
```{r download-data}
if(!file.exists("UCI HAR Dataset")) {
    if(!file.exists("dataset.zip")){
        fileURL <- "http://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip"
        download.file (fileURL, "dataset.zip", mode = "wb")
    }
    unzip("dataset.zip")
}
```

The data files are unarchived into _"UCI HAR Dataset"_ folder, and ready to be loaded. According to the _README.txt_ file from the data set, and the assignment description, the following 8 data files should be loaded into R environment for processing. Several messages are printed to the console to notify users the data-loading progress. They are commented out here.
```{r load-data, cache=TRUE}
# message("Loading data ...")
activityLabels <- read.table("UCI HAR Dataset/activity_labels.txt")
features <- read.table("UCI HAR Dataset/features.txt")
# message("25% ...")
test_x <- read.table("UCI HAR Dataset/test/X_test.txt")
test_y <- read.table("UCI HAR Dataset/test/y_test.txt")
# message("50% ...")
test_subject <- read.table("UCI HAR Dataset/test/subject_test.txt")
train_x <- read.table("UCI HAR Dataset/train/X_train.txt")
# message("75% ...")
train_y <- read.table("UCI HAR Dataset/train/y_train.txt")
train_subject <- read.table("UCI HAR Dataset/train/subject_train.txt")
# message("100% Done!")
```

### Exploratory Analysis

According to _README.txt_, _activityLabels_ contains numeric values to their corresponding types of activites.
```{r activity-labels}
summary(activityLabels)
```

_features_ contains the numeric values to their corresponding measurements.
```{r features}
summary(features)
```

_test\_subject_ and _train\_subject_ contain the corresponding subject numbers for each observation from both training set and test set.
```{r subject}
table(test_subject); table(train_subject)
```

_test\_y_ and _train\_y_ contain the corresponding activity numbers for each observation from both training set and test set.
```{r activity}
table(test_y); table(train_y)
```

_test\_x_ and _train\_x_ contain all the measurement data for each subject for each activities.
```{r mearement-data}
dim(test_x); dim(train_x)
```

### Step 1: Merge training and test sets

Based on previous analysis and the dimensions of the data frames, training set and test set can be merged using _rbind()_ function.
```{r merge-data}
data <- rbind(test_x, train_x)
activity <- rbind(test_y, train_y)
subject <- rbind(test_subject, train_subject)
```

_data_ is the merged measurements data frame, and their corresponding activiy and subject numbers are merged into _activity_ and _subject_ data frames respectively.
```{r combined-data}
table(activity); table(subject)
```

### Step 2: Extract mean() and std() mearuement data

According to _features\_info.txt_ file, and my personal understanding of the assignment, only those measurements matches exactly to _mean()_ or _std()_ are selected using _grep()_ function. Measurements such as _fBodyAcc-meanFreq()-X_ and _fBodyAcc-meanFreq()-X_ are discarded.
```{r extract-data}
meanIdx <- grep("mean\\(\\)", features$V2)
stdIdx <- grep("std\\(\\)", features$V2)
data <- data[, c(meanIdx, stdIdx)]
dim(data)
```

_meanIdx_ and _stdIdx_ are the index of those mearuements labels matches _mean()_ and _std()_, respectively. They will be used again later when applying column names to the tidied data set.

### Step 3: Use descriptive activity names

Using _factor()_ function, with activity names extracted from _activityLabels_ as factor labels, _activity_ is converted to a factor with corresponding types of activities.
```{r activity-label}
activity$V1 <- factor(activity$V1, labels = activityLabels$V2)
table(activity)
```

### Step 4: Label column names

First, combine _subject_, _activity_, and _data_ into one data frame.
```{r combine-data}
tidyData <- cbind(subject, activity, data)
```

Get column labels using _meanIdx_ and _stdIdx_, replacing "-" and "()" in the measurement names with "\_" so that the column names from the output data can be correctly read into R using _read.table()_.
```{r column-label}
meanLabel <- gsub("__", "", gsub('[^A-Za-z]', '_', features$V2[meanIdx]))
stdLabel <- gsub("__", "", gsub('[^A-Za-z]', '_', features$V2[stdIdx]))
names(tidyData) <- c("Subject", "Activity", as.character(meanLabel), as.character(stdLabel))
names(tidyData)
```

R working environment can be cleaned up with only _tidyData_ data frame left.
```{r clean-up}
# Remove all other unecessary data frames and variables
rm(list=setdiff(ls(), "tidyData"))
```

The integrity of _tidyData_ can be verified whether each subject has performed each types of activity.
```{r data-test}
table(tidyData[,1:2])
```

### Step 5: Output data

A second, independent tidy data set is created, with the average of each variable for each activity and each subject. First, using _dplyr::group\_by()_ function group _tidyData_ by _Subject_ and _Activity_. Then using _dplyr::summarise\_each()_ function apply _mean()_ to each column of the data frame. Then write the resulting data frame to _output.txt_ file, with row names neglected.
```{r output-data}
output <- group_by(tidyData, Subject, Activity) %>% summarise_each(funs(mean))
write.table(output, file="output.txt", row.names = FALSE)
```
